---
title: "Directed Acyclic Graphs (DAGs) in R"
subtitle: "Workplace physical and social factors on low back pain"
author: "Kenta Okuyama, MPH, PhD"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  revealjs::revealjs_presentation:
    theme: simple
    highlight: tango
    transition: fade
    center: false
    incremental: true
    slide_level: 2
    self_contained: true
bibliography: references.bib
# csl: apa.csl
link-citations: false
css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  out.width = "75%",
  fig.width = 6.3,
  fig.height = 3.6
)

library(ggdag)
library(dagitty)
```


## Agenda (40 min)

- Why DAGs in epidemiology (5 min)
- Study example (5 min): workplace physical and social factors → low back pain [@tamekuni2026]
- Hands-on: build a DAG in R (25-30 min)
- Q&A

## Why DAGs?

DAGs help you:

- make causal assumptions explicit
- identify confounders
- decide what to (and what not to) adjust for
- communicate analysis decisions transparently

```{r}

```

<small> Core references for causal diagrams in epidemiology: [@greenland1999; @shrier2008; @hernan2020] </small>

## Key terms - confounder
- Common cause of exposure & outcome
```{r}

```

## Key terms - mediator
- Lies on the causal pathway

```{r}

```

## Key terms - collider
- Common effect of two variables
```{r}

```

## Key terms - effect modifier
- Modify the effect of exposure on outcome
```{r}

```

## Adjustment rules (for total effect)
- adjust confounders
- avoid adjusting mediators
- avoid adjusting/conditioning on colliders

<small> [@greenland1999; @hernan2020] </small>

## Why “adjust for confounders”?

```{r}

```

## Why not “adjust for mediators”?

```{r}

```

## Why not “adjust for colliders”?

```{r}

```

## Building a DAG

Study example:

- Cross-sectional survey-based study
- Japanese workers at small/medium sized enterprises (n=)
- Workplace physical and social factors → Low back pain

<small> [@tamekuni2026] </small>

## Background and hypothesis

- Previous studies on occupational pain investigated physical and social factors but without causal structure
- High physical load (physical factor) → Low back pain
- Good interpersonal relations (social factor) may mitigate the association

## Variables

- Physical load (exposure)
- Low back pain (outcome)
- Age, Gender, Education, Smoking, Drinking, Body mass index (BMI), Work hours, Job statisfaction, Depression, Interpersonal relations stress (confounders/mediators/effect modifiers)

## Start with exposure → outcome

```{r}
library(ggdag)
# 1. Name and position variables in a x-y dimension (physical load = p_load, low back pain = lbp)
coords1 <- list(
  x = c(p_load = 1, lbp = 3),
  y = c(p_load = 0, lbp = 0)
)

# 2. Define casual structures
g1 <- dagify(
  lbp ~ p_load,
  
  # set position of variables
  coords = coords1,
  
  # label variables
  labels = c(
    "lbp" = "LB pain",
    "p_load" = "Physical\n load"  
    ),
  
  # define exposure and outcome
  exposure = "p_load",
  outcome = "lbp"
)
```

## DAG 1
```{r}
ggdag(g1, text = FALSE, use_labels = "label") +
  theme_dag()
```

## Add socio-demographic factors
```{r}
# 1. Name and position variables in a x-y dimension
coords2 <- list(
  x = c(age = -2.5, gender = -2.5, education = -2, income = -2, p_load = 1, lbp = 3),
  y = c(age = 0, gender = -0.5, education = 0.5, income = 1, p_load = 0, lbp = 0)
)

# 2. Define casual structures
g2 <- dagify(
  lbp ~ p_load + age + gender + education + income,
  p_load ~ age + gender + education,
  education ~ gender,
  income ~ age + gender + education,
  coords = coords2,
  labels = c(
    "lbp" = "LB pain",
    "p_load" = "Physical\n load",
    "age" = "Age",
    "gender" = "Gender",
    "education" = "Education",
    "income" = "Income"
  ),
  exposure = "p_load",
  outcome = "lbp"
)
```

## DAG 2
```{r}
ggdag(g2, text = FALSE, use_labels = "label") +
  theme_dag()
```




## Add lifestyle factors

```{r}
# 1. Name and position variables in a x-y dimension
coords3 <- list(
  x = c(age = -2, gender = -2, education = -2, income = -2, drink = -1, smoke = -1, sleep = -1, pa = -1, p_load = 1, lbp = 3),
  y = c(age = 0, gender = -0.5, education = 0.5, income = 1, drink = 1, smoke = 0.5, sleep = -0.5, pa = -1, p_load = 0, lbp = 0)
)

# 2. Define causal structures
g3 <- dagify(
  lbp ~ p_load + age + gender + education + income + drink + smoke + sleep + pa,
  p_load ~ age + gender + education,
  education ~ gender,
  income ~ age + gender + education,
  drink ~ age + gender + education + income + pa,
  smoke ~ age + gender + education + income + pa,
  sleep ~  age + gender + drink + smoke + pa + p_load + lbp,
  pa ~ age + gender + drink + smoke + sleep + p_load + lbp,
  
  coords = coords3,
  labels = c(
    "lbp" = "LB pain",
    "p_load" = "Physical\n load",
    "age" = "Age",
    "gender" = "Gender",
    "education" = "Education",
    "drink" = "Drinking",
    "smoke" = "Smoking",
    "sleep" = "Sleep\n quality",
    "pa" = "Physical\n activity",
    "income" = "Income"
  ),
  exposure = "p_load",
  outcome = "lbp"

)
```

## DAG 3
```{r}
ggdag(g3, text = FALSE, use_labels = "label") +
  theme_dag()
```


## Step 3: Introduce PhysicalLoad as a potential effect modifier

We now ask:

Does interpersonal stress have a different effect on chronic LBP when physical load is high vs low?

Effect modification is a modeling/interpretation question; DAGs mainly help with confounding control. [@hernan2020]

But we still must decide: is PhysicalLoad also a confounder?

We’ll show two defensible options.

## Option A: PhysicalLoad is an effect modifier only (not a confounder)

Assumption:

- Physical load affects chronic LBP,
- but does not cause interpersonal relations stress (after accounting for job type, etc.)

## Option A: PhysicalLoad is an effect modifier only (not a confounder)
```{r}
dag3A <- dagitty("dag {
  InterpersonalStress -> LBP_Chronic
  InterpersonalStress -> MentalHealth -> LBP_Chronic

  PhysicalLoad -> LBP_Chronic

  Age -> InterpersonalStress
  Age -> MentalHealth
  Age -> LBP_Chronic
  Age -> PhysicalLoad

  Sex -> InterpersonalStress
  Sex -> MentalHealth
  Sex -> LBP_Chronic
  Sex -> PhysicalLoad

  JobType -> InterpersonalStress
  JobType -> LBP_Chronic
  JobType -> PhysicalLoad

  WorkHours -> InterpersonalStress
  WorkHours -> MentalHealth
  WorkHours -> LBP_Chronic

  PriorLBP -> InterpersonalStress
  PriorLBP -> LBP_Chronic
}")

```

## Option A: PhysicalLoad is an effect modifier only (not a confounder)
```{r}
ggdag(dag3A, layout = "stress") +
  geom_dag_edges() +
  geom_dag_label(size = 3.3) +
  theme_dag()
```

## Option B: PhysicalLoad is both a confounder and effect modifier

Assumption:
- Physical load influences interpersonal stress (e.g., demanding physical roles affect teamwork/strain),
- and also affects chronic LBP.

## Option B: PhysicalLoad is both a confounder and effect modifier
```{r}
dag3B <- dagitty("dag {
  InterpersonalStress -> LBP_Chronic
  InterpersonalStress -> MentalHealth -> LBP_Chronic

  PhysicalLoad -> InterpersonalStress
  PhysicalLoad -> LBP_Chronic

  Age -> InterpersonalStress
  Age -> MentalHealth
  Age -> LBP_Chronic
  Age -> PhysicalLoad

  Sex -> InterpersonalStress
  Sex -> MentalHealth
  Sex -> LBP_Chronic
  Sex -> PhysicalLoad

  JobType -> InterpersonalStress
  JobType -> LBP_Chronic
  JobType -> PhysicalLoad

  WorkHours -> InterpersonalStress
  WorkHours -> MentalHealth
  WorkHours -> LBP_Chronic

  PriorLBP -> InterpersonalStress
  PriorLBP -> LBP_Chronic
}")

```

## Option B: PhysicalLoad is both a confounder and effect modifier
```{r}
ggdag(dag3B, layout = "stress") +
  geom_dag_edges() +
  geom_dag_label(size = 3.3) +
  theme_dag()
```

## Compute minimal adjustment sets (Option A)

```{r}
adjustmentSets(dag3A, exposure = "InterpersonalStress", outcome = "LBP_Chronic")
```

Interpretation:

- Under Option A, physical load may not be required for confounding control.
- You can still include it to test effect modification.

<small> dagitty reference: [@textor2016] </small>


## Compute minimal adjustment sets (Option B)

```{r}
adjustmentSets(dag3B, exposure = "InterpersonalStress", outcome = "LBP_Chronic")
```

Interpretation:

- Under Option B, physical load is part of confounding control and can be used for effect modification.

## Practical decision slide (what you say)

- Pick one DAG that matches your substantive belief.

- Choose one feasible adjustment set.

- Explain why you did not adjust for:
  - MentalHealth (mediator) if total effect
  - any selection variables / colliders (if applicable)

<small> [@greenland1999; @shrier2008; @hernan2020] </small>



## Paper-style modeling approach (illustrative)

Your paper used multinomial logistic regression for none / acute / chronic. [@tamekuni2026]

Below is a template showing:

- baseline adjustment (choose based on your DAG)
- effect modification test via interaction term

```{r eval=FALSE, include=FALSE}
# install.packages("nnet")
library(nnet)

# Example data requirements
# df$LBP_cat <- factor(df$LBP_cat, levels = c("none", "acute", "chronic"))
# Variables: InterpersonalStress, PhysicalLoad, Age, Sex, JobType, WorkHours, PriorLBP

# Choose an adjustment set (example):
# adj_vars <- c("Age","Sex","JobType","WorkHours","PriorLBP")  # replace as needed

# Fit model with effect modification:
fit_em <- multinom(
  LBP_cat ~ InterpersonalStress * PhysicalLoad + Age + Sex + JobType + WorkHours + PriorLBP,
  data = df
)

summary(fit_em)

# Chronic vs none coefficients:
coef(fit_em)["chronic", ]
```


## How to present effect modification (cleanly)

Instead of only reporting an interaction p-value, show predicted chronic LBP probabilities (or ORs)
across interpersonal stress, by physical load strata.

<small> (Once you know your variable coding, you can add a plot chunk here.) </small>

## Take-home points

- DAGs make causal assumptions explicit and improve adjustment decisions. [@greenland1999]
- Separate confounders vs mediators vs colliders to avoid bias. [@shrier2008; @hernan2020]
- dagitty makes the “why we adjusted for X” reproducible. [@textor2016]
- Physical load can be:
  - effect modifier only (Option A), or
  - both confounder and effect modifier (Option B)

## References

::: {#refs}
:::

