---
title: "Directed Acyclic Graphs (DAGs) in R"
subtitle: "Workplace physical and social factors on low back pain"
author: "Kenta Okuyama, MPH, PhD"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  revealjs::revealjs_presentation:
    theme: simple
    highlight: tango
    transition: fade
    center: false
    incremental: true
    slide_level: 2
    self_contained: true
    css: styles.css
bibliography: references.bib
link-citations: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  out.width = "75%",
  fig.width = 6.3,
  fig.height = 3.6
)
library(ggdag)
```


## Agenda (40 min)

- Introduction of DAGs (5 min)
- Study example (5 min): workplace physical and social factors → low back pain [@tamekuni2026]
- Hands-on: build a DAG in R (25-30 min)
- Q&A

## Introduction of DAGs

DAGs help you:

- make causal assumptions explicit
- identify confounders
- decide what to (and what not to) adjust for
- communicate analysis decisions transparently

<small> Core references for causal diagrams in epidemiology: [@greenland1999; @shrier2008; @hernan2020] </small>

## Key terms - confounder
- Common cause of exposure & outcome

```{r echo=FALSE}
coords1 <- list(
  x = c(exposure = 1, outcome = 3),
  y = c(exposure = 0, outcome = 0)
)
g1 <- dagify(
  outcome ~ exposure,
  coords = coords1,
  labels = c(
    "exposure" = "Exposure",
    "outcome" = "Outcome"  
    ),
  exposure = "exposure",
  outcome = "outcome"
)
ggdag(g1, text = FALSE, use_labels = "label") +
  theme_dag()
```

## Key terms - mediator
- Lies on the causal pathway

```{r, echo=FALSE}
coords2 <- list(
  x = c(exposure = 1, mediator = 2, outcome = 3),
  y = c(exposure = 0, mediator = 0.5, outcome = 0)
)
g2 <- dagify(
  outcome ~ exposure + mediator,
  exposure ~ mediator,
  coords = coords2,
  labels = c(
    "exposure" = "Exposure",
    "outcome" = "Outcome",
    "mediator" = "Mediator"
    ),
  exposure = "exposure",
  outcome = "outcome"
)
ggdag(g2, text = FALSE, use_labels = "label") +
  theme_dag()
```

## Key terms - collider
- Common effect of two variables

```{r, echo=FALSE}
coords3 <- list(
  x = c(exposure = 1, mediator = 2, collider = 2, outcome = 3),
  y = c(exposure = 0, mediator = 0.5, collider = -0.5, outcome = 0)
)
g3 <- dagify(
  outcome ~ exposure + mediator,
  exposure ~ mediator,
  collider ~ exposure + outcome,
  coords = coords3,
  labels = c(
    "exposure" = "Exposure",
    "outcome" = "Outcome",
    "mediator" = "Mediator",
    "collider" = "Collider"
    ),
  exposure = "exposure",
  outcome = "outcome"
)
ggdag(g3, text = FALSE, use_labels = "label") +
  theme_dag()
```

## Key terms - effect modifier

- Modify the effect of exposure on outcome

```{r, echo=FALSE}
coords4 <- list(
  x = c(exposure = 1, mediator = 2, collider = 2, emodifier = 2.5, outcome = 3),
  y = c(exposure = 0, mediator = 0.5, collider = -0.5, emodifier = -1, outcome = 0)
)
g4 <- dagify(
  outcome ~ exposure + mediator + emodifier,
  exposure ~ mediator,
  collider ~ exposure + outcome,
  coords = coords4,
  labels = c(
    "exposure" = "Exposure",
    "outcome" = "Outcome",
    "mediator" = "Mediator",
    "collider" = "Collider",
    "emodifier" = "Effect modifier"
    ),
  exposure = "exposure",
  outcome = "outcome"
)
ggdag(g4, text = FALSE, use_labels = "label") +
  theme_dag()
```

## Adjustment rules (for total effect)
- adjust confounders
- avoid adjusting mediators
- avoid adjusting/conditioning on colliders

<small> [@greenland1999; @hernan2020] </small>

## Why “adjust for confounders”?

```{r}

```

## Why not “adjust for mediators”?

```{r}

```

## Why not “adjust for colliders”?

```{r}

```

## Building a DAG

Study example:

- Cross-sectional survey-based study
- Japanese workers at small/medium sized enterprises (n=)
- Workplace physical and social factors → Low back pain

<small> [@tamekuni2026] </small>

## Background and hypothesis

- Previous studies on occupational pain investigated physical and social factors but without causal structure
- High physical load (physical factor) → Low back pain
- Good interpersonal relations (social factor) may mitigate the association

## Variables

- Physical load (exposure)
- Low back pain (outcome)
- Age, Gender, Education, Smoking, Drinking, Body mass index (BMI), Job statisfaction, Depression, Interpersonal relations stress (confounders/mediators/effect modifiers)

## Start with exposure → outcome

```{r}
library(ggdag)
# 1. Name and position variables in a x-y dimension (physical load = p_load, low back pain = lbp)
coords1 <- list(
  x = c(p_load = 1, lbp = 3),
  y = c(p_load = 0, lbp = 0)
)

# 2. Define casual structures
g1 <- dagify(
  lbp ~ p_load,
  
  # set position of variables
  coords = coords1,
  
  # label variables
  labels = c(
    "lbp" = "LB pain",
    "p_load" = "Physical\n load"  
    ),
  
  # define exposure and outcome
  exposure = "p_load",
  outcome = "lbp"
)
```

## DAG 1
```{r}
ggdag(g1, text = FALSE, use_labels = "label") +
  theme_dag()
```

## Add socio-demographic factors
```{r}
# 1. Name and position variables in a x-y dimension
coords2 <- list(
  x = c(age = -2.5, gender = -2.5, education = -2, income = -2, p_load = 1, lbp = 3),
  y = c(age = -0.25, gender = -1, education = 0.5, income = 1, p_load = 0, lbp = 0)
)

# 2. Define casual structures
g2 <- dagify(
  lbp ~ p_load + age + gender + education + income,
  p_load ~ age + gender + education,
  education ~ gender,
  income ~ age + gender + education,
  coords = coords2,
  labels = c(
    "lbp" = "LB pain",
    "p_load" = "Physical\n load",
    "age" = "Age",
    "gender" = "Gender",
    "education" = "Education",
    "income" = "Income"
  ),
  exposure = "p_load",
  outcome = "lbp"
)
```

## DAG 2
```{r}
ggdag(g2, text = FALSE, use_labels = "label") +
  theme_dag()
```

## Add lifestyle factors

```{r}
# 1. Name and position variables in a x-y dimension
coords3 <- list(
  x = c(age = -2.5, gender = -2.5, education = -2, income = -2, drink = -1, smoke = -1, sleep = -0.5, pa = -1, p_load = 1, lbp = 3),
  y = c(age = -0.25, gender = -1, education = 0.5, income = 1, drink = 1, smoke = 0.5, sleep = -0.5, pa = -1, p_load = 0, lbp = 0)
)

# 2. Define causal structures
g3 <- dagify(
  lbp ~ p_load + age + gender + education + income + drink + smoke + sleep + pa,
  p_load ~ age + gender + education,
  education ~ gender,
  income ~ age + gender + education,
  drink ~ age + gender + education + income,
  smoke ~ age + gender + education + income,
  sleep ~  age + gender + education + income + drink + smoke + pa,
  pa ~ age + gender + education + income,

  coords = coords3,
  labels = c(
    "lbp" = "LB pain",
    "p_load" = "Physical\n load",
    "age" = "Age",
    "gender" = "Gender",
    "education" = "Education",
    "drink" = "Drinking",
    "smoke" = "Smoking",
    "sleep" = "Sleep\n quality",
    "pa" = "Physical\n activity",
    "income" = "Income"
  ),
  exposure = "p_load",
  outcome = "lbp"
)
```

## DAG 3
```{r}
ggdag(g3, text = FALSE, use_labels = "label") +
  theme_dag()
```

## Add the remaining variables

```{r}
# 1. Name and position variables in a x-y dimension
coords4 <- list(
  x = c(age = -2.5, gender = -2, education = -2, income = -2, drink = -1, smoke = -1, sleep = -0.5, pa = -1, bmi = 0, job_sat = 2, depress = 2, int_stress = 1, p_load = 1, lbp = 3),
  y = c(age = -0.25, gender = -1, education = 0.5, income = 1, drink = 1, smoke = 0.5, sleep = -0.5, pa = -1, bmi = -1.25, job_sat = -1, depress = 1, int_stress = -0.75, p_load = 0, lbp = 0)
)

# 2. Define causal structures
g4 <- dagify(
  lbp ~ p_load + age + gender + education + income + drink + smoke + sleep + pa + bmi + job_sat + depress + int_stress,
  p_load ~ age + gender + education,
  education ~ gender,
  income ~ age + gender + education,
  drink ~ age + gender + education + income,
  smoke ~ age + gender + education + income,
  sleep ~  age + gender + education + income + drink + smoke + pa,
  pa ~ age + gender + education + income,
  bmi ~ age + gender + education + income + drink + smoke + sleep + pa,
  job_sat ~ p_load + int_stress,
  depress ~ p_load + int_stress,
  int_stress ~ age + gender + education,
  
  coords = coords4,
  labels = c(
    "lbp" = "LB pain",
    "p_load" = "Physical\n load",
    "age" = "Age",
    "gender" = "Gender",
    "education" = "Education",
    "drink" = "Drinking",
    "smoke" = "Smoking",
    "sleep" = "Sleep\n quality",
    "pa" = "Physical\n activity",
    "income" = "Income",
    "bmi" = "BMI",
    "job_sat" = "Job\n satisfaction",
    "depress" = "Depression",
    "int_stress" = "Interpersonal\n stress"
  ),
  exposure = "p_load",
  outcome = "lbp"

)
```

## DAG 4
```{r}
ggdag(g4, text = FALSE, use_labels = "label") +
  theme_dag()
```


## Identifying an adjustment set

```{r}
# dagitty::adjustmentSets(g4, exposure = "p_load", outcome = "lbp")
```

Interpretation:

- 
- 

<small> dagitty reference: [@textor2016] </small>


## Practical decision slide (what you say)

- Pick one DAG that matches your substantive belief.

- Choose one feasible adjustment set.

- Explain why you did not adjust for:
  - MentalHealth (mediator) if total effect
  - any selection variables / colliders (if applicable)

<small> [@greenland1999; @shrier2008; @hernan2020] </small>

## Take-home points

- DAGs make causal assumptions explicit and improve adjustment decisions. [@greenland1999]
- Separate confounders vs mediators vs colliders to avoid bias. [@shrier2008; @hernan2020]
- dagitty makes the “why we adjusted for X” reproducible. [@textor2016]
- Physical load can be:
  - effect modifier only (Option A), or
  - both confounder and effect modifier (Option B)

## References {.refs-slide}

::: {#refs}
:::
